Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-18 _tex/index.tex
Built site for gh-pages

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to exhibit a higher trait level. For example, when assessing text
quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
the other hand, research on validity suggests that scores generated by
CJ can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

@@

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
resulting from comparing two stimuli is determined by two factors: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects variability in the stimulus' perceived trait
level. Figure~\ref{fig-discriminal_process} shows example distributions
of discriminal process for two stimuli (objects).

@@

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
As a result, its assumptions may not suit applications beyond the
psychophysical contexts for which it was created. Thurstone himself
cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
assumptions could prove problematic with more complex, less homogeneous
stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
applications often deal with complex, less homogeneous stimuli, two key
assumptions of Case V may not consistently hold in theory or practice:
the zero correlation and equal dispersion between stimuli.

{The \emph{zero correlation between stimuli} represents two additional
assumptions about the method:}

@@

Building on the previous section, it is evident that the BTL model
typically functions as the measurement model for the trait of interest
\citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying texts' quality level (the
latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

>>>>

In \emph{comparative judgment} (CJ) studies, judges assess a specific
trait or attribute across various stimuli by performing pairwise
comparisons \citep{Thurstone_1927a, Thurstone_1927b}. Each comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to exhibit a higher trait level. For example, when assessing text
quality, judges compare pairs of written texts (the stimuli) to
determine the relative quality each text exhibit (the trait)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
the other hand, research on validity suggests that scores generated by
CJ can accurately represent the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

@@

In its most general form, Thurstone's theory
\citeyearpar{Thurstone_1927b} posits that the dichotomous outcome
resulting from comparing two stimuli is determined by two factors: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} refers to the psychological effect each
stimulus has on the judges, or more simply stated, the judges'
perception of the trait level of each stimulus. Thurstone assumes that
the discriminal process for each stimulus follows a Normal distribution.
In this distribution, the mode (mean), known as the \emph{modal
discriminal process}, represents the position of the stimulus on the
trait continuum, while the dispersion, known as the \emph{discriminal
dispersion}, reflects the variability in the stimulus' perceived trait
level. Figure~\ref{fig-discriminal_process} shows example distributions
of discriminal process for two stimuli (objects).

@@

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
As a result, its assumptions may not suit applications beyond the
psycho-physical contexts for which it was created. Thurstone himself
cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
assumptions could prove problematic with more complex, less homogeneous
stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
applications often deal with complex, less homogeneous stimuli, two key
assumptions of Case V may not consistently hold in theory or practice:
the zero correlation and equal dispersion between stimuli.

The assumption of \emph{zero correlation between stimuli} is best
illustrated with an example. For instance, when evaluating text quality,
this assumption suggests that a judge's perception of quality in one
text does not influence their perception of the same trait in the
comparison text. Thurstone attributes this independence to the
cancellation of potential judges' biases, driven by opposing and equally
weighted factors; commonly known as `mood' and `simultaneous contrast'
effects, that operate during pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. This cancellation of bias has been
mathematically demonstrated by \citet{Andrich_1978}, assuming additive
biases and a logit scale derived from the BTL model.

However, given that evidence has shown judges' quality assessments are
influenced by multiple, complex aspects of the texts
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}, it is
plausible that the accuracy of these judgments also varies based on
factors such as age, culture, education, expertise
\citep[pp.~683]{Kelly_et_al_2022}, and ultimately, based on the
individual judge themselves
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020}.

{This means that the cancellation of judges' biases may not hold in
theory or practice, and that judges' biases may depend on the stimuli,
by means of an interaction (but present) or because the judges works as
a confounder}

Such differences may not be detected through analyses of bias and misfit
\citep[pp.~683]{Kelly_et_al_2022}.

@@

Building on the previous section, it is evident that the BTL model
commonly functions as the measurement model for the trait of interest in
CJ experiments \citep{Andrich_1978, Bramley_2008}. A measurement model
specifies how manifest variables contribute to the estimation of latent
variables \citep{Everitt_et_al_2010}. For example, when evaluating text
quality, researchers use the BTL model to process the dichotomous
outcomes resulting from the pairwise comparisons (the manifest
variables) to estimate scores that reflect the underlying quality level
of texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
